NumPy---Array Programming in Python

### 2020-01-30 body.tex
Revise section title
Fix name
Edit ecosystem

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It is a community developed, open source library, which provides a
multidimensional Python array object along with array-aware functions
that operate on it.
Because of its inherent simplicity, the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for array programming.

@@

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs, or
narrative computational notebooks.  This rich and productive
environment has made the scientific Python ecosystem hugely popular
for scientific research.

Despite its tight integration, the ecosystem was conceived to have
clear functional separation, with well defined interfaces.  This makes
possible the design of large, complex scientific libraries, that
add specialized functionality for solving specific problems.

@@

\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights some of the essential tools and libraries in the vast
   scientific Python ecosystem.  Most of the ecosystem builds on NumPy to
   provide a rich and powerful collection of array-aware functionality.
  }
  \label{fig:ecosystem}
\end{figure}

For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image 
reconstruction.
SciPy is used to provide tools used in general image processing such as 
filtering and image alignment, while scikit-image---a popular image processing
library that extends SciPy---provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling general optimization tasks.
NetworkX \cite{SciPyProceedings_11}---a popular package for complex
network analysis---is used for analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including the generation of the final image of the black hole.

Similar examples can be found in physics, earth science, biology,
and more.
% FIXME: this should be expanded 
Over the last few years there has been a massive increase in the use
of scientific Python due in part to the rapid growth of data science,
machine learning, and artificial intelligence.  
As a result, NumPy and its API has become ubiquitous.

\section*{Proliferation of arrays}

>>>>

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It is a community developed, open source library, which provides a
multidimensional Python array object along with array-aware functions
that operate on it.
Because of its inherent simplicity, the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become widespread as
a language for array programming.

@@

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs, or
narrative computational notebooks.  This rich and productive
environment has made Python popular
for scientific research.

@@

\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights just a few of the essential tools and libraries in the vast
   scientific Python ecosystem.  Most of the ecosystem builds on NumPy arrays to
   provide a rich and powerful collection of array-aware functionality.
  }
  \label{fig:ecosystem}
\end{figure}

Many research groups have designed large,
complex scientific libraries, that add specialized functionality for
solving problems in specific domains.
For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration for radio interferometry imaging, analysis, and simulation
relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image 
reconstruction.
SciPy is used to provide tools used in general image processing such as 
filtering and image alignment, while scikit-image, a popular image processing
library that extends SciPy, provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling general optimization tasks.
NetworkX \cite{SciPyProceedings_11}, a popular package for complex
network analysis, is used for analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including the generation of the final image of the black hole.

Similar examples can be found in physics, earth science, biology,
and more.
In many fields, the scientific Python ecosystem is the primary
software environment.
The recent rapid growth of data science, machine learning, and
artificial intelligence has dramatically increased scientific
Python use.
NumPy and its ecosystem is now taught in
numerous university courses, bootcamps, and summer schools.
There are multiple community conferences and workshops.
NumPy and its API has become ubiquitous.

\section*{Array proliferation and interoperability}
